---
title: "Parse PSMs"
author:
  - name: "Tom Smith"
    affiliation: "Cambridge Centre for Proteomics"
date: "`r format(Sys.time(), '%d %B, %Y')`"
abstract: | 
  Here, we parse the PSM-level PD output and QC
output:
  pdf_document:
  html_notebook: default
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
---

Load libraries
```{r setup, message=FALSE}

#### Load packages ####
library(camprotR)
library(ggplot2)
library(tidyverse)
library(MSnbase)
```

Read in PSM data
```{r}
infiles <- Sys.glob('../raw/total*PSMs.txt.gz')
names(infiles) <- c('AGC: 2E5',  'AGC: 5E4')
psm <- infiles %>% lapply(read.delim)
print(infiles)
```

Make the cRAP list for filtering
```{r}
get_fasta_ids <- function(fasta){
  # Load the FASTA 
  bs.fasta <- Biostrings::fasta.index(fasta, seqtype = "AA")
  
  # Extract the UniProt accessions 
  accessions <- bs.fasta %>% 
    pull(desc) %>% 
    stringr::str_extract_all("(?<=\\|).*?(?=\\|)") %>% 
    unlist()

  accessions
}

crap.accessions <- get_fasta_ids('../../reanalysis_published/shared_files/cRAP_FullIdentifiers.fasta')


```

Match species to uniprotID
```{r}
hs.accessions <- get_fasta_ids(
  '../shared_files/CCP_SwissProt_homo_sapiens_proteome_without_isoforms_20180409.fasta.gz')
sc.accessions <- get_fasta_ids(
  '../shared_files/swissprot_yeast_database_june2020.fasta.gz')

uniprot_2_species <- data.frame('id'=c(hs.accessions, sc.accessions),
                                'species'=c(rep('H.sapiens', length(hs.accessions)),
                                            rep('S.cerevisiae', length(sc.accessions))))
head(uniprot_2_species)
```

Parse and filter PSMs to remove cRAP proteins
```{r}
psm_parsed <- psm %>% lapply(function(x){
  parse_features(x, TMT=TRUE, level='PSM',
                 crap_proteins=crap.accessions, unique_master=FALSE)
})
```


Annotated the data with the species
```{r}
psm_parsed_annt <- psm_parsed %>% lapply(function(x){
  
 species_matches <- x %>% select(Protein.Accessions) %>%
  mutate(Protein.Accessions_sep=Protein.Accessions) %>%
  separate_rows(Protein.Accessions_sep) %>%
  merge(uniprot_2_species, by.x='Protein.Accessions_sep', by.y='id', all.x=TRUE) %>%
  group_by(Protein.Accessions) %>%
  summarise(all_species=paste0(unique(species), collapse='; ')) %>%
  mutate(species=ifelse(grepl(';', all_species), 'mixed', all_species))
 
 x %>% merge(species_matches, by='Protein.Accessions')
})
```
Checking the above hasn't altered `nrow`
```{r}
psm_parsed %>% lapply(nrow)
psm_parsed_annt %>% lapply(nrow)
```

Summarise PSMs per species
```{r}
psm_parsed_annt %>% names() %>% lapply(function(x){
  p1 <- psm_parsed_annt[[x]] %>%
    group_by(species) %>%
    tally() %>%
    ggplot(aes(species, n)) +
    geom_bar(stat='identity') +
    theme_camprot() +
    xlab('') +
    ylab('PSMs') +
    ggtitle(x)
  
  p2 <- psm_parsed_annt[[x]] %>%
    select(Master.Protein.Accessions, species) %>%
    unique() %>%
    group_by(species) %>% tally() %>%
    ggplot(aes(species, n)) +
    geom_bar(stat='identity') +
    theme_camprot() +
    xlab('') +
    ylab('Proteins') +
    ggtitle(x)
  
  print(p1)
  print(p2)
  
  print(psm_parsed_annt[[x]] %>%
    select(Master.Protein.Accessions, species) %>%
    unique() %>%
    group_by(species) %>% tally())
  
  ggsave(sprintf('../results/plots/%s_psm_n.png', gsub('AGC: ', '', x)), p1)
  ggsave(sprintf('../results/plots/%s_proteins_n.png', gsub('AGC: ', '', x)), p2)
  
  return(NULL)
})
```
Checking the number of proteins in each, and the intersection
```{r}
length(unique(psm_parsed_annt$`AGC: 2E5`$Master.Protein.Accessions))
length(unique(psm_parsed_annt$`AGC: 5E4`$Master.Protein.Accessions))
length(intersect(unique(psm_parsed_annt$`AGC: 2E5`$Master.Protein.Accessions),
                 unique(psm_parsed_annt$`AGC: 5E4`$Master.Protein.Accessions)))
```


Make MSnSets
```{r}

sample_info <- read.delim('../raw/sample_info.tsv') %>% tibble::column_to_rownames('tag')

psm_res <- psm_parsed_annt %>% lapply(function(x){
  # Abundance columns for TMT PD-output start with Abundance 
  abundance_cols <- colnames(x)[grepl('Abundance.', colnames(x))]
  
  .e <- as.matrix(x[,abundance_cols])
  .f <- x[,setdiff(colnames(x), abundance_cols)]
  
  # update the column names to remove the 'Abundance.` prefix
  colnames(.e) <- gsub('Abundance.', '', colnames(.e))
  
  res <- MSnbase::MSnSet(exprs=.e, fData=.f, pData=sample_info)
  
  res
})
```

Plotting the distribution of tag intensities in each full dataset and the single species subsets. Note that the tag intensities for yeast fall into the 3 groups we expect given the experimental design.
```{r}

psm_res %>% names() %>% lapply(function(x){
  
  all <- psm_res[[x]]
  hs <- all[fData(all)$species=='H.sapiens']
  sc <- all[fData(all)$species=='S.cerevisiae']
  
  slices <- list('All'=all, 'H.sapiens'=hs, 'S.cerevisiae'=sc)
  for(slice in names(slices)){
    p <- slices[[slice]] %>% log(base=2) %>% plot_quant() +
      ggtitle(sprintf('%s - %s', x, slice)) +
      ylab('PSM intensity (log2)')
    print(p)
    
    p <- slices[[slice]] %>% log(base=2) %>% plot_quant(method='density') +
      xlab('PSM intensity (log2)') +
      ggtitle(sprintf('%s - %s', x, slice))
    print(p)
  }
  return(NULL)
})

```




Save for downstream notebooks
```{r}
saveRDS(psm_res, '../results/psm_res.rds')
```



