---
title: 'Aggregate to protein'
author:
  - name: "Tom Smith"
    affiliation: "Cambridge Centre for Proteomics"
date: "`r format(Sys.time(), '%d %B, %Y')`"
abstract: | 
  Here, aggregate the PSM-level abundances to peptide-level and protein-level
  abundances and re-inspect the fold changes
output:
  pdf_document:
  html_notebook: default
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
---

Load libraries
```{r setup, message=FALSE}

#### Load packages ####
library(camprotR)
library(ggplot2)
library(tidyverse)
library(MSnbase)
```

Read in PSM-level quant and expected fold changes
```{r}
psm_res_flt <- readRDS('../results/psm_res_flt.rds')
expected <- readRDS('../results/expected.rds')
```

Function to aggregate PSMs to peptide or protein-level abundances
```{r}
aggregate_psms <- function(psm, level='peptide'){
  
  if(level=='peptide'){
    feature_name <- 'Sequence'
    min_features <- 1
  } else{
    feature_name <- 'Master.Protein.Accessions'
    min_features <- 2
  }
  
  psm_filt <- psm %>% 
    MSnbase::filterNA(pNA = 0) %>% # remove PSMs with missing values 
    restrict_features_per_protein(min_features = min_features, # remove proteins with <2 PSMs
                                  master_protein_col=feature_name, plot=FALSE) 
  
  group <- MSnbase::fData(psm_filt)[[feature_name]]

  agg <- psm_filt %>% MSnbase::combineFeatures(groupBy = group, method = 'sum')

  feature_counts <- count_features_per_protein(psm_filt)
  
  return(list('agg'=agg, 'feature_counts'=feature_counts))
}
```


Aggregate to protein-level abundance
```{r}

prot_res <- psm_res_flt %>% lapply(function(agc_target){
  agc_target %>% lapply(function(thresholded_psms){
  
    # log center-median normalise PSMs before aggregation
    x <- thresholded_psms %>% log(base=2) %>%
      MSnbase::normalise(method='diff.median')
    exprs(x) <- 2^exprs(x) # then exponentiate to get untransformed intensities
    
    aggregate_psms(x, level='protein')
  })
})

```




Aggregate to peptide-level abundance
```{r}

pep_res <- psm_res_flt %>% lapply(function(agc_target){
  
  agc_target %>% lapply(function(thresholded_psms){
  
    # log center-median normalise PSMs before aggregation
    x <- thresholded_psms %>% log(base=2) %>%
      MSnbase::normalise(method='diff.median')
    exprs(x) <- 2^exprs(x) # then exponentiate to get untransformed intensities
    
    aggregate_psms(x, level='peptide')
  })
    
})




```


How many proteins per AGC target and threshold parameters are retained.
```{r}

  
plot_n_features <- function(obj, feature_name='Proteins'){
  p <- obj %>%
    lapply(function(x) lapply(x, function(y) nrow(y$agg))) %>% unlist() %>%
    data.frame() %>% setNames('n_proteins') %>%
    tibble::rownames_to_column('threshold') %>%
    separate(threshold, into=c('agc', 'interference', 'mean_sn', 'notch'), sep='\\.|, ') %>%
    mutate(notch=recode(notch, '1'='With notch filt', '0'='Without notch filt.')) %>%
    ggplot(aes(interference, mean_sn, fill=n_proteins, label=n_proteins)) +
    geom_tile() +
    theme_camprot(base_size=15) +
    facet_grid(notch~agc) +
    geom_text() +
    scale_fill_gradient(low='grey90', high=get_cat_palette(1), name=feature_name) +
    xlab('Maximum inteference (%)') +
    ylab('Minimum mean signal/noise')
  
  print(p)
}

prot_res %>% plot_n_features()
pep_res %>% plot_n_features('Peptides')
```
```{r}
saveRDS(prot_res, '../results/prot_res.rds')
saveRDS(pep_res, '../results/pep_res.rds')
```

*Can re-run cells below without re-aggregating if required*
```{r}
prot_res <- readRDS('../results/prot_res.rds')
pep_res <- readRDS('../results/pep_res.rds')
```

Get fold changes for protein and peptide-level abundances.
```{r}
source('../R/get_quant_vs_mean.R')
```

Plot protein-level tag vs group fold changes
```{r}
summarisations <- list('protein'=prot_res, 'peptide'=pep_res)
```

```{r}
# define thresholds to plot fold changes. Each set of threshold will be plotted at peptide and protein level, with both AGC settings
datasets <- c('100, 0, 0') 
```


```{r}
summarisations %>% names() %>% lapply(function(level){
  
  summarisations[[level]] %>% names() %>% lapply(function(agc){
    
    datasets %>% lapply(function(thresholds){
        
      if(level=='protein'){
        feature_name <- 'Master.Protein.Accessions'
      } else{
        feature_name <- 'Sequence'
      }
      
      x <- get_notch_per_protein(psm_res_flt[[agc]][[thresholds]],
                                 master_prot_col=feature_name) %>%
        mutate(sample=remove_x(sample))
      y <-  get_quant_vs_mean(summarisations[[level]][[agc]][[thresholds]]$agg)
      
      to_plot <- merge(x, y,
                       by.x=c(feature_name, 'sample'),
                       by.y=c('id', 'tag')) %>%
        filter(species!='mixed')
      
      p <- to_plot  %>%
        ggplot(aes(log2(intensity), diff)) +
          theme_camprot(base_size=12) +
          facet_grid(species~comparison) +
          geom_hline(aes(yintercept=log2(expected)),
                     data=expected,
                     colour='black', linetype=2) +
          xlab(sprintf('%s abundance (log2)', level)) +
          ylab('Difference in abundnace (log2)') +
          ggtitle(sprintf('%s - %s - %s', agc, thresholds, level))
    
      suppressWarnings(print(p + geom_point(size=0.2, alpha=0.1, colour='grey30')))
    
      suppressWarnings(print(p %+% to_plot[to_plot$fraction_below>0,] +
              aes(fraction_below) +
              geom_point(size=0.5, alpha=0.5) +
               geom_smooth() +
              xlab('Fraction PSMs below notch')))
      return(NULL)
      
    })
    
    return(NULL)
    
  
  })
  
  return(NULL)
  
})
```





